\documentclass[a4paper,14pt]{extarticle}
\usepackage{color}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{bbold}
\usepackage{mathtools}
\usepackage{centernot}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{caption}
\usepackage{adjustbox}
\usepackage{multicol}  

\theoremstyle{definition}
\newtheorem*{theorem}{Theorem}
\newtheorem*{definition}{Definition}
\newtheorem*{lemma}{Lemma}
\newtheorem*{corollary}{Corollary}
\newtheorem*{proposition}{Proposition}
\newtheorem*{eg}{Example}
\newtheorem*{remark}{Remark}


\begin{document}

\title{\textbf{Topology and Groups - MATH0074}}
\author{\textbf{Based on lectures by Dr. Lars Louder}\\ Notes taken by Imran Radzi}
\date{}
\maketitle

\pagenumbering{roman}
(Revision) notes based on the Autumn 2021 Topology and Groups lectures by Dr. Lars Louder.
Some parts marked with (*) (+) are taken from Hatcher's Algebraic Topology and 
the site on Topology and Groups by Prof. Jonny Evans.
\begingroup
\let\cleardoublepage\clearpage
\tableofcontents
\endgroup
\newpage
\pagenumbering{arabic}

\section{Point-set Topology}

\subsection{Preliminaries}
\begin{definition}[Topological space]
  A topological space is a pair $(X,\mathcal{T})$ such that
  \begin{enumerate}
    \item $X$ is a set 
    \item $\mathcal{T}\subset \mathcal{P}(X)$ is a collection of subsets of $X$
    \item $\emptyset\in\mathcal{T}, \,X\in\mathcal{T}$
    \item $\mathcal{T}$ is closed under finite intersections and arbitrary unions
  \end{enumerate}
\end{definition}

\begin{definition}[Open neighbourhood]
  If $x\in X, \,U$ open in $X$, and $x\in U$, then $U$ is an \emph{open neighbourhood} of $x$.
\end{definition}

\begin{definition}[Hausdorff spaces]
  A topological space $(X,\mathcal{T})$ is \emph{Hausdorff} if $\forall \,x,y\in X$, there 
  exists $U,V$ open neighbourhoods of $x,y$ respectively such that $U\cap V=\emptyset$.
\end{definition}

\begin{definition}[Homeomorphisms]
  A map $f:X\rightarrow Y$ is a \emph{homeomorphism} if 
  \begin{enumerate}
    \item $f$ is bijective
    \item $f$ is continuous
    \item $f^{-1}$ is continuous
  \end{enumerate}
\end{definition}

\begin{definition}[Continuous maps]
  A map $f:X\rightarrow Y$ is continuous if $\forall \,U\text{ (open)}\subset Y$,
  $f^{-1}(U)$ is open in $X$.
\end{definition}

\begin{definition}
  If $\mathcal{T}$ and $\mathcal{T}'$ are topologies on $X$ such that $\mathcal{T}\subsetneq\mathcal{T}'$ then $\mathcal{T}'$ is \emph{finer} than $\mathcal{T}$, and 
  $\mathcal{T}$ is \emph{coarser} than $\mathcal{T}'$.
\end{definition}

\begin{proposition}
  $\text{id}:(X,\mathcal{T})\rightarrow (X,\mathcal{T}')$ is continuous if and only if 
  $\mathcal{T}$ is finer than $\mathcal{T}'$.
\end{proposition}

\begin{definition}[Subspace topology]
  If $X$ is a topological space, $Y\subset X$, the subspace topology on $Y$ is defined by
  \[U\text{ open in } Y\iff \exists V\text{ open in }X \text{ such that }U=Y\cap V\]
\end{definition}

\begin{definition}
  If a map $f:X\rightarrow Y$ is continuous, the \emph{image} of $f$ is the set 
  \[f(X)=\{f(x)\,|\,x\in X\}\subset Y\] with the subspace topology.
\end{definition}

\begin{definition}[Product topology]
  Let $X, Y$ be spaces. The \emph{product topology} on $X\times Y$ is the smallest (coarsest)
  topology making the projections
  \[p_X:X\times Y\rightarrow X, \,\,\,p_Y:X\times Y\rightarrow Y\]
  continuous.
\end{definition}

\begin{proposition}
  Product of Hausdorff spaces is Hausdorff.
\end{proposition}

\subsection{Connectedness}
\begin{definition}[Connectedness]
  A space $X$ is \emph{disconnected} if there exists a surjective continuous map 
  $f:X\rightarrow\{p_1,p_2\}$. A space is \emph{connected} if every continuous function
  $f:X\rightarrow\{p_1,p_2\}$ is constant.
\end{definition}

\begin{definition}
  A pair of sets $U,V\subset X$ is said to disconnect $X$ if they are non-empty, disjoint,
  $U\cup V=X$ and both are open.
\end{definition}

\begin{definition}
  $X$ is disconnected if there exists $U,V$ which disconnect $X$.
\end{definition}


\begin{definition}[Path]
  A \emph{path} in $X$ is a continuous map $\gamma:[0,1]\rightarrow X$. $\gamma$ is a path 
  from $\gamma(0)$ to $\gamma(1)$. $a,b\in X$ are said to be connected by a path if there 
  is a path from $a$ to $b$.
\end{definition}

\begin{definition}[Path-connectedness]
  A space $X$ is \emph{path-connected} if for all $x,y$, there exists
  \[\gamma:[0,1]\rightarrow X\text{ such that }\gamma(0)=x, \,\gamma(1)=y\]
\end{definition}

\noindent or equivalently,

\begin{definition}
  We say $X$ is path-connected if there exists a unique equivalence class, where the 
  equivalence relation $\sim$ is defined $a\sim b$ if and only if there exists a path from 
  $a$ to $b$.
\end{definition}

\begin{proposition}
  Suppose $X$ is connected. Then, if $f:X\rightarrow Y$, then $f(X)\subset Y$ is connected.
\end{proposition}

\begin{proposition}
  $[0,1]$ is connected.
\end{proposition}

\begin{corollary}
  If $X$ is path-connected, then $X$ is connected.
\end{corollary}

\begin{definition} $X\subset\mathbb{R}$ is an \emph{interval} if $a\leq b\leq c$, 
  $a,c\in X\implies b\in X$. 
\end{definition}

\begin{proposition}
  A subset of $\mathbb{R}$ is connected if and only if it is an interval.
\end{proposition}

\begin{definition}[Locally (path) connected]
  A space $X$ is locally (path) connected at a point $p$ if for every open neighbourhood $U$ of $p$, 
  there exists a (path) connected open neighbourhood $V$ of $p$ such that $p\in V\subset U$.
\end{definition}

\begin{proposition}
  If $X$ is locally path-connected then the path components of $X$ are open.  
\end{proposition}

\begin{proposition}
  If $X$ is connected and locally path-connected, then $X$ is path connected.
\end{proposition}


\subsection{Compactness}
\begin{definition}[Open cover]
  An \emph{open cover} of a space $X$ is a collection of open sets $\mathcal{U}$ such that
  \[X=\bigcup_{U\in\mathcal{U}}U\]
\end{definition}

\begin{definition}
  A space $X$ is \emph{compact} if every open cover has a finite subcover.
\end{definition}

\begin{lemma}
  Closed subsets of compact spaces are compact.
\end{lemma}

\begin{theorem}
  If $X,Y$ are compact, then $X\times Y$ is compact.
\end{theorem}

\begin{theorem}[Heine-Borel theorem]
  $X\subset\mathbb{R}^n$ is compact if and only if $X$ is closed and bounded.
\end{theorem}

\begin{theorem}
  $[0,1]$ is compact.
\end{theorem}

\begin{theorem}
  If $f:X\rightarrow Y$ is continuous, $X$ compact, then $f(X)\subset Y$ is compact with respect to the subspace topology.
\end{theorem}

\begin{proposition}
  If $C\subset Y$ is compact, $Y$ Hausdorff, then $C$ is closed.
\end{proposition}

\begin{proposition}
  If $f:X\rightarrow Y$ is a continuous bijection, $X$ compact, $Y$ Hausdorff, then $f$ 
  is a homeomorphism
\end{proposition}

\subsection{Quotient spaces}
\begin{definition}[Quotient map]
  Let $q:X\rightarrow Y$ be a continuous surjection. Then $q$ is a \emph{quotient map}
  if $q^{-1}(Y)$ is open if and only if $U$ is open. (A bijective quotient map is a 
  homeomorphism)
\end{definition}

\begin{definition}[Quotient space]
  Let $X$ be a space, and $\sim$ an equivalence relation on $X$, and 
  $q:X\rightarrow X/\sim = Y$ the quotient map. The quotient topology on $Y$ is defined by
  $U$ open in $Y$ if and only if $q^{-1}(U)$ is open in $X$.
\end{definition}

\begin{lemma}
  \[
  \begin{tikzcd}
    X \arrow{r}{f} \arrow[swap]{dr}{q} & Z  \\
     & Y \arrow{u}{h}
  \end{tikzcd}
\]
  Let $f$ be continuous, and suppose $f$ factors through $q:X\rightarrow Y$, a quotient 
  map, i.e., $\exists h:Y\rightarrow Z$ such that $h\circ q = f$. Then $h$ is continuous.
\end{lemma}

\begin{proposition}
  Let $f:X\rightarrow Y$ be a continuous surjection with $X$ compact, $Y$ Hausdorff. Then 
  $f$ is a quotient map. 
\end{proposition}

\begin{definition}[Disjoint union]
  Let $X_1,X_2$ be topological spaces. The \emph{disjoint union} of $X_1$ and $X_2$,
  $X_1\sqcup X_2$, is the space with the underlying set $X_1\sqcup X_2$, with $U$ open 
  in $X_1\sqcup X_2$ if and only if $U\cap X_1$ is open in $X_1$, and $U\cap X_2$ is 
  open in $X_2$.
\end{definition}

\begin{definition}[Cell complex]
  A \emph{cell complex} is a space built up inductively, as follows
  \begin{enumerate}
    \item ($n=0$) We start with a discrete set $X^{(0)}$ consisting of points, which we call 
    $0$-cells $\{e_i^0\,|\,i\in I_0\}, \,e_i^0\cong pt$. $X^{(0)}=\bigsqcup_i e_i^0$ is called the $0$-skeleton.
    \item ($n>0$) We add a (possibly empty) subset of $n$-cells $\{e_i^n\,|\,i\in I_n\}$ 
    $e_i^n\cong D^n$, the $n$-dimensional disk, and a continuous map 
    \[\phi_i^n:\partial e_i^n\cong S^{n-1}\rightarrow X^{(n-1)}\] and here the $n$-skeleton 
    is \[X^{(n)}=X^{(n-1)}\sqcup\bigsqcup e_i^n/\sim\]
  \end{enumerate}
  A space $X$ is a cell complex if there exists $X^{(0)}\subset X^{(1)}\subset\ldots$ 
  as above, with the condition that $U$ is open in $X$ if and only if 
  $X^{(n)}\cap U$ is open for all $n$. \\

  $X^{(0)}\subseteq X^{(1)}\subseteq\ldots$ is called the \emph{cell decomposition} of $X$.
\end{definition}

\begin{definition}
  The suspension $SX$ of a space $X$ is the space
  \[SX=X\times I/\sim\] where $(x,t)\sim (x',t')$ if and only if $(x,t)=(x',t')$ or 
  $t=t'=1$ or $t=t'=0$.
\end{definition}

\begin{proposition}
  $SS^n$ is homeomorphic to $S^{n+1}$. $SD^n$ is homeomorphic to $D^{n+1}$.
\end{proposition}

\begin{definition}[Presentation complex]
  Given a group $G$ and its presentation, the presentation complex of $G$ with respect
  to the given presentation is the 
  $2$-dimension cell complex with 1 vertex, obtained by attaching a loop (1-cell) at the vertex 
  for each generator of $G$, and attaching a 2-cell along every relation
  in the presentation, where the boundary of the 2-cell is attached according to 
  the appropriate word.
\end{definition}

\begin{definition}[Cayley graph]
  Given a group $G$ and its presentation, and $S$ a (possibly generating) set of $G$,
  then the Cayley graph $C(G,S)=(G\sqcup G\times I\times S)/\sim$ where the equivalence
  relation $\sim$ is given by $g\sim(g,0,s)$, $gs\sim(g,1,s)$.
\end{definition}

\section{Homotopy}

\subsection{Homotopy}

\begin{definition}
  Let $(X,A)$ be a pair of spaces, where $A\subseteq X$, $f_0,f_1:X\rightarrow Y$. We say
  $f_0$ and $f_1$ are \emph{homotopic relative} to $A$ if there exists a continuous 
  function $F$ where
  $F:X\times I\rightarrow Y$ such that $F(-,0)=f_0, \,F(-,1)=f_1$ and $F(a,t)=f_0(a)=f_1(a)$ 
  for all $t$. In this case we write $f_0\simeq_A f_1$. \\

  If $A=\emptyset$ then we say $f_0$ and $f_1$ are \emph{homotopic} and write $f_0\simeq f_1$.
\end{definition}

\begin{lemma}[*]
  A function defined on the union of two closed sets is continuous if it is continuous when 
  restricted to each of the closed sets separately.
\end{lemma}

\begin{proposition}
  Any two continuous maps $f_0,f_1:X\rightarrow \mathbb{R}^n$ are homotopic via the homotopy
  \[F(x,t)=tf_1(x)+(1-t)f_0(x)\]
\end{proposition}

\begin{definition}[Homotopy equivalence]
  Two spaces $X$ and $Y$ are \emph{homotopy equivalent} if there exists 
  $f:X\rightarrow Y$, $g:Y\rightarrow X$ such that 
  $f\circ g\simeq\text{id}_Y$, $g\circ f\simeq\text{id}_X$. In this case, we write 
  $X\simeq Y$.
\end{definition}

\begin{proposition}
  Homotopy equivalence is an equivalence relation on (topological) spaces.
\end{proposition}

\begin{proposition}
  $\mathbb{R}^n\simeq pt$
\end{proposition}

\begin{definition}
  A space $X$ is \emph{contractible} if $X\simeq pt$, or in other words,
  $\text{id}:X\rightarrow X$ is homotopic to 
  a constant map. In this case the map $\text{id}_X$ is said to be \emph{null-homotopic}.
\end{definition}

\begin{proposition}
  $\mathbb{R}^n\setminus pt\simeq S^{n-1}$
\end{proposition}

\begin{proposition}
  If $f:X\rightarrow S^2$ is a non-surjective map then $f$ is homotopic to a constant 
  map.
\end{proposition}

\begin{definition}
  The cone $CX$ on a space $X$ is the space
  \[CX=X\times I/\sim\] where $(x,t)\sim(x',t')$ if and only if $(x,t)=(x',t')$ or 
  $t=t'=1$.
\end{definition}

\begin{proposition}
  $CX$ is always contractible.
\end{proposition}

\begin{proposition}
  If $X$ is contractible then $X$ is path-connected.
\end{proposition}

\begin{definition}[Retract]
  Let $A\subseteq X$ be a subspace. $A$ is a \emph{retract} of $X$ if there exists a continuous map $r:X\rightarrow A$ (retraction) such that $r|_A=\text{id}_A$. $A$ is a \emph{deformation
  retract} of $X$ if there exists such a function $r$ such that $r$ is homotopic to 
  $\text{id}_X$ relative to $A$.
\end{definition}

\begin{proposition}
  If $A$ is a deformation retract of $X$ then $X\simeq A$.
\end{proposition}

\begin{definition}[Homotopy extension property]
  A pair of spaces $(X,A)$ has the \textit{homotopy extension property} if for any 
  $f:X\rightarrow Y$, and homotopy $H$ of $h=f|_A$, this homotopy extends to a homotopy 
  $F$ of $f$.
\end{definition}

\begin{theorem}
  If $A\subseteq X$, $X$ a cell-complex and $A$ a subcomplex, then $(X,A)$ has the 
  homotopy extension property.
\end{theorem}

\begin{corollary}
  If $A\subseteq X$, $X$ a CW complex, $A$ a subcomplex, $A$ contractible, then the 
  quotient map $q:X\rightarrow X/A$ (collapse $A$ to a single point) is a homotopy 
  equivalence.
\end{corollary}

\subsection{Paths and path-homotopy}

\begin{definition}[Path-homotopy]
  Two paths $\gamma_0$ and $\gamma_1$ are \emph{path-homotopic} if they are homotopic 
  relative to $\{0,1\}\subseteq I$. In particular $\gamma_0(0)=\gamma_1(0)$, 
  $\gamma_0(1)=\gamma_1(1)$. If $F$ is a homotopy from $\gamma_0$ to $\gamma_1$,
  \[F(-,0)=\gamma_0(0), \,F(-,1)=\gamma_1(1)\] $F$ is a family of paths connecting 
  $\gamma_0(0)$ and $\gamma_0(1)$
\end{definition}

\begin{proposition}
  Path-homotopy is an equivalence relation on the set of paths in (a topological space) $X$.
\end{proposition}

\begin{definition}[Based loop]
  A \emph{based loop} at $x_0\in X$ is a path $\gamma:I\rightarrow X$ such that 
  $\gamma(0)=\gamma(1)=x_0$.
\end{definition}

\begin{definition}[Fundamental group of a space]
  The \emph{fundamental group} of $X$ at $x_0$ is the set (group)
  \[\{[\gamma]\,|\,\gamma\text{ is a loop based at }x_0\}\]
  which is denoted by $\pi_1(X,x_0)$.
\end{definition}

\begin{definition}[$n^{\text{th}}$ homotopy group]
  The $n^{\text{th}}$ homotopy group of a space $X$ at $x_0$ is the set (group)
  \[\pi_n(X,x_0)=\{[f:I^n\rightarrow X\,|\,f(\partial I^n)\rightarrow x_0]\}\]
\end{definition}

\begin{definition}
  A loop based at $x_0$ is null-homotopic if it is path-homotopic to a constant path.
\end{definition}

\begin{definition}[Free homotopy]
  If $\gamma_0$ and $\gamma_1$ are based loops (not necessarily at the same point), then 
  $\gamma_0$ and $\gamma_1$ are \emph{freely homotopic} if they are homotopic through 
  based loops, so if $F$ is a free homotopy between $\gamma_0$ and $\gamma_1$, then,
  \[F(x_0)=\gamma_0, \,F(x,1)=\gamma_1\] \[F(0,t)=F(1,t)\text{ for all }t\]
\end{definition}

\begin{proposition}
  Free homotopy is an equivalence relation on the set of based loops in (a topological space)
  $X$.
\end{proposition}

\begin{definition}
  A based loop \emph{bounds a disk} if the induced map \[\bar{\gamma}:[0,1]/
  {\scriptstyle 0=1}\cong S^1
  \subseteq D^2\] extends to a continuous function $D^2\rightarrow X$.
\end{definition}

\begin{lemma}
  The following are equivalent
  \begin{enumerate}
    \item $\gamma$ bounds a disk. 
    \item $\gamma$ is null-homotopic.
    \item $\gamma$ is freely homotopic to a constant path.
  \end{enumerate}
\end{lemma}

\section{Covering spaces}
\begin{definition}[Covering map]
  A map $p:X'\rightarrow X$ is a \emph{covering map} if $\forall x\in X$, there exists
  $U$, an open neighbourhood of $x$, and a discrete set $\Delta$ and a homeomorphism 
  $h_U:U\times\Delta\rightarrow p^{-1}(U)$ such that \[p\circ h_u=\pi_U:U\times\Delta\rightarrow 
  U\] and such a neighbourhood $U$ is called a \emph{covering neighbourhood}.
\end{definition}

\begin{definition}[Lift]
  Let $f:Y\rightarrow X$ and $g:Z\rightarrow X$ be two maps,
  \[
    \begin{tikzcd}
      &Z\arrow{d}{g} \\ Y\arrow{ur}{\tilde{f}} \arrow{r}{f}&X
    \end{tikzcd}\]
  a lift of $f$ is a map $\tilde{f}:Y\rightarrow Z$ such that \[g\circ\tilde{f}=f\]
\end{definition}

\subsection{Path/Homotopy lifting lemma}
\begin{lemma}[Path/Homotopy lifting lemma]
  Let $p:X'\rightarrow X$ be a covering map and $f:I^n\rightarrow X$ a continuous map.
  Then for any $x'\in p^{-1}(f(U))$, there exists a unique lift $\tilde{f}$ of $f$ to 
  $X'$, where $\tilde{f}(0)=x'$. 
\end{lemma}

\begin{definition}
  A covering space $p:X'\rightarrow X$ is \emph{trivial} if $X$ is a covering neighbourhood.
\end{definition}

\begin{lemma}
  Suppose $p:X'\rightarrow X$ is a trivial covering map, and $f:Y\rightarrow X$ is continuous, 
  $Y$ connected, the for any $y_0\in Y$ and $x'\in p^{-1}(f(y_0))$, there exists a unique 
  lift $\tilde{f}:Y\rightarrow X'$ such that $\tilde{f}(y_0)=x'$.
\end{lemma}

\begin{lemma}
  Let $X$ be a compact metric space. Then a continuous function $f:X\rightarrow\mathbb{R}$  attains a maximum and minimum value on $X$.
\end{lemma}

\begin{lemma}[Lebesgue's number lemma]
  Let $X$ be a compact metric space, $\mathcal{U}$ an open cover of $X$, then there 
  exists $\epsilon>0$ such that for all $x\in X$, there exists $U\in\mathcal{U}$ such that 
  $B_\epsilon(x)\subseteq U$. \\
  Such an $\epsilon$ is called the Lebesgue number for $\mathcal{U}$.
\end{lemma}

\begin{lemma}[(+)]
  Let $p:X'\rightarrow X$ be a covering space, and $f:Y\rightarrow X$ a continuous map, $Y$ 
  connected. Then two lifts $\tilde{f_1},\tilde{f_2}:Y\rightarrow X'$ are equal for all 
  $y\in Y$ if and only if they are equal for some $y\in Y$.
\end{lemma}

\begin{corollary}
  If $[\gamma]\in\pi_1(X,x_0)$ and there exists a covering space $X'$ of $X$ so that 
  $\gamma$ lifts to a non-closed path then $[\gamma]\neq1\in\pi_1(X,x_0)$.
\end{corollary}

\begin{corollary}
  \[\pi_1(S^1)\neq1\]
\end{corollary}

\begin{corollary}
  $\text{id}_{S^1}:S^1\rightarrow S^1$ is \emph{not} null-homotopic. In particular $S^1$
  is not contractible.
\end{corollary}

\subsection{Winding numbers}
\begin{definition}
  Let $\gamma$ be a closed path in $S^1$. The \emph{winding number} of $\gamma$, 
  $\omega(\gamma)$ is the integer $\tilde{\gamma(1)}-\tilde{\gamma(0)}$ where $\tilde{\gamma}$
  is any lift of $\gamma$ to $\mathbb{R}$.
\end{definition}

\begin{proposition}
  $\omega(\gamma)$ is well-defined, and only depends on the free homotopy classes of $\gamma$.
\end{proposition}

\begin{proposition}
  If $\gamma\simeq\gamma'$ (freely homotopic) then $\omega(\gamma)=\omega(\gamma')$.
\end{proposition}

\subsection{Covering transformations}

\begin{definition}[Covering transformation]
  Let $p:X'\rightarrow X$ be a covering map. A \emph{covering transformation} is a homeomorphism $h:X'\rightarrow X'$ such that $p\circ h=p$
\end{definition}

\begin{theorem}
  If $X'$ is the universal cover of space $X$, then 
  \[\pi_1(X,x_0)=\{h:X'\rightarrow X'\,|\,p\circ h =p\}\]
\end{theorem}

\section{More on fundamental groups and covering spaces}
\begin{definition}
  Let $\alpha, \beta$ be paths in a space $X$. We say $\alpha, \beta$ are \textit{composable}
  if \[\alpha(1)=\beta(0)\text{ (note that the order matters)}\]
\end{definition}

\noindent If $\alpha,\beta$ are composable, their product $\alpha\cdot\beta$ is the path 
\[\alpha\cdot\beta=\begin{cases}
  \alpha(2t) & 0\leq t\leq \frac{1}{2} \\
  \beta(2t-1) & \frac{1}{2}\leq 1
\end{cases}\]
i.e., we traverse through $\alpha, \beta$ with twice the speed. Note that in some sources
they may write $\beta\cdot\alpha$ instead of $\alpha\cdot\beta$ to mean the same thing.

\begin{definition}
  If $\alpha$ is a path (not necessarily closed), then $\bar{\alpha}$ is the 
  pathdefined by $\bar{\alpha}=\alpha(1-t)$, i.e., traversing through $\alpha$ in the 
  backwards direction. 
\end{definition}

\begin{theorem}
  Let $X$ be a \textit{path-connected} space, $x_0$ a basepoint. The set \[\pi(X,x_0)=\{\text{loops based at }x_0\}/\text{path homotopy}\] with multiplication given by $[\alpha]\cdot[\beta]=[\alpha\cdot\beta]$, for any $[\alpha],[\beta]\in
  \pi(X,x_0)$, and inverses given by $[\alpha]^{-1}=[\bar{\alpha}]$,
  and the identity given by $1=[x_0]$, defines a group, called the \textit{fundamental group}
  (of $X$).
\end{theorem}
We note that it makes sense to talk about \textit{the} fundamental group, if we restrict 
the spaces in question to path-connected spaces, so we shall restrict our attention to 
path-connected spaces from now on.

\begin{theorem}
  Let $\alpha$ be a path from $x_0$ to $x_1$. The map \[[\alpha]_*:\pi_1(X,x_0)\rightarrow \pi_1(X,x_1)\] defined by 
  \[[\gamma]\mapsto[\bar{\alpha}\cdot\gamma\cdot\alpha]\] is an isomorphism.
\end{theorem}

\begin{theorem}
  Let $(X,x_0)$, $(Y,y_0)$ be two (pointed) path-connected spaces, and 
  $f:(X,x_0)\rightarrow (Y,y_0)$ be continuous and such that $f(x_0)=y_0$. Then this map 
  induces the map 
  \[f_*:\pi_1(X,x_0)\rightarrow \pi_1(Y,y_0)\] defined by $f_*([\gamma])=[f\circ\gamma]$.
\end{theorem}

\begin{theorem}
  With conditions as the previous theorem, along with another path-connected space 
  $(Z,z_0)$ and another continuous $g:(Y,y_0)\rightarrow (Z,z_0)$, we have 
  \[(g\circ f)_* = g_*\circ f_*\]
\end{theorem}

In other words, the previous two theorems show that $\pi$ is a functor taking 
the category of topological spaces to the category of groups.

\begin{theorem}
  Let $f_t:X\times I\rightarrow Y$ be a homotopy of $f_0$ and $f_1$, and let $\alpha$ 
  be the path $f_t(x_0)$ from $f_0(x_0)=y_0$ to $f_1(x_0)=y_1$. Then,
  \[[\alpha]_*\circ (f_0)_*=(f_1)_*\]
\end{theorem}

\begin{corollary}
  Let $X, Y$ be path-connected and homotopy equivalent. Then,
  \[\pi_1(X)\cong\pi_1(Y)\]
\end{corollary}

\begin{theorem}[Brouwer's no retraction theorem]
  There does not exist $r:D^2\rightarrow S^1$ such that $r|_{S^1}=\text{id}_{S^1}$.
\end{theorem}

\begin{theorem}[Brouwer's fixed point theorem]
  Let $f:D^2\rightarrow D^2$ be continuous. Then $f$ has a fixed point, i.e.,
  \[\exists x\in D^2\text{ such that }f(x)=x\]
\end{theorem}

\subsection{Classification of covering spaces}
\begin{definition}
  Let $Y_0, Y_1$ be two covers of a space $X$. We say that $Y_0$ and $Y_1$ are \textit{equivalent} if there exists a homeomorphism $h:Y_0\rightarrow Y_1$ such that 
  \[p_1\circ h=p_0\] (and $h(y_0)=y_1$ if $Y_0, Y_1$ have base points). \\

  If $Y_0=Y_1$, an equivalence $h:Y_0\rightarrow Y_0$ is a \textit{covering transformation}.
  It also follows that 
  \[Aut(Y)=\{h:Y\rightarrow Y \,|\, p\circ h=p\}\]
\end{definition}

\begin{lemma}
  Let $p:(Y,y_0)\rightarrow(X,x_0)$ be a cover. Then 
  \[p_*:\pi_1(Y,y_0)\rightarrow\pi_1(X,x_0)\]
  is injective.
\end{lemma}
This way we get a map from the pointed covering spaces of $(X,x_0)$ (quotient according 
to equivalences) to the subgroups of $\pi_1(X,x_0)$, where $X$ is path-connected and 
locally contractible.

\begin{proposition}
  $p_*(\pi_1(Y,y_0))<\pi_1(X,x_0)$
\end{proposition}

\begin{theorem}[Classification of covering spaces]
  Let $X$ be path-connected and locally contractible.
  \begin{enumerate}
    \item The map from the set of connected covering spaces under the quotient of equivalence,
    \[\{(Y,y_0)\}/\text{equiv}\rightarrow\{\text{subgroups of }\pi_1(X,x_0)\}\}\] defined by
    \[(Y,y_0)\mapsto p_*(\pi(Y,y_0))\] is \textit{bijective}.
    \item Given (1), $H<\pi_1(X,x_0)$ let $(X_H,x_H)$ be the 
    cover of $\pi_1(X,x_0)$ corresponding to $H$. $H<K\iff\exists(X_H,x_H)\xrightarrow{h}
    (X_K,x_K)$
    \begin{enumerate}
      \item[i.] $h$ is a cover
      \item[ii.] $[H:K]=\#h^{-1}(X_K)$
    \end{enumerate}
    \item \[Aut(X_H)=\{h:X_H\rightarrow X_H\,|h\text{ a covering transformation}\}\] Then
    \[X_H\cong N(H)/H\] In particular if $H=1<\pi_1(X,x_0)$, then
    \[Aut(X_H)\cong\pi_1(X,x_0)\] in which case $X_H$ is the \textit{universal cover}.
  \end{enumerate}
\end{theorem}

\begin{proposition}
  Define $f:\pi_1(X,x_0)\rightarrow p^{-1}(x)$ by \[[\gamma]\mapsto\tilde{\gamma}(1)\] Then,
  \[f([\gamma_1])=f([\gamma_2])\iff [\gamma_1][\gamma_2]^{-1}\in H, \,[\gamma_1]\in H[\gamma_2]
  \] i.e., $[\gamma_1],[\gamma_2]$ determine the same right coset of $H$, and the map 
  $f$, factors through $H\backslash G$ (through the map $\tilde{f}$)
  \[
  \begin{tikzcd}
    G \arrow{r}{f} \arrow[swap]{dr}{} & p^{-1}(x_0) \\
     & H\backslash G \arrow{u}{\tilde{f}}
  \end{tikzcd}
\]
and by a similar argument, we get a bijection 
\[\text{f}\rightarrow p^{-1}(x_0), \,\,\,\tilde{f}(H)=y_0\]
\end{proposition}

\begin{corollary}
  \[[H:G]=|H\backslash G|=|p^{-1}(x_0)|\] i.e., degree of the cover is equal to the index 
  of the subgroup.
\end{corollary}

\begin{theorem}[General lifting lemma]
  \[
  \begin{tikzcd}
    & (Y,y_0) \arrow{d}{p}  \\
    (Z,z_0) \arrow{r}{f} \arrow{ur}{\tilde{f}}  &  (X,x_0)
  \end{tikzcd}
\]
Suppose $Z$ is path-connected and locally contractible. Then there exists a unique lift 
\[\tilde{f}:(Z,z_0)\rightarrow(Y,y_0)\iff f_*(\pi_1(Z,z_0))\subset p_*(\pi_1(Y,y_0))\]
\end{theorem}

\begin{corollary}
  \[
  \begin{tikzcd}
     (Y,y_0) \arrow[rr, shift left, "h_0"] \arrow{dr}{p_0} && (Y_1,y_1) \arrow[ll, shift left, "h_1"]{h_1}
     \arrow{dl}{p_1}  \\
     & (X,x_0)
  \end{tikzcd}
\]
Suppose $p_{0_*}(\pi_1(Y_0,y_0))=p_{1_*}(\pi_1(Y_1,y_1))$, then,
\[\exists!h:(Y_0,y_0)\xrightarrow{\cong}(Y_1,y_1)\]
so any two path-connected locally contractible pointed covers which correspond 
to the same subgroup are equivalent. In particular this proves the injectivity claim
in statement (1) of the theorem for classification of covering spaces.
\end{corollary}

\begin{theorem} Let $G=\pi_1(X,x_0), \,g\in G$, and $H=p_*(\pi_1(X_H,x_H))$, so 
  $g^{-1}Hg=p_*(\pi_1(X_H,x_H\cdot g))$.
  \[
    \begin{tikzcd}
       (X_H,x_H) \arrow[rr, "h"] \arrow{dr}{p} && (X_H,x_H\cdot g) 
       \arrow{dl}{p}  \\
       & (X,x_0)
    \end{tikzcd}
  \]
  By the general lifting lemma, we know $\exists h\iff H<g^{-1}Hg$. If such an $h$ exists,
  then it is also a covering transformation (isomorphism), so equivalently we have 
  \[\#h^{-1}(x_H\cdot g)=1\] i.e., $g^{-1}Hg=H$, which is equivalent to $g\in N(H)$. Hence 
  we have, the following, 
  \[\exists h_g:(X_H,x_H)\xrightarrow{\cong}(X_H,x_H\cdot g)\iff g\in N(H)\]
  which also defines a surjective map 
  \[N(H)\rightarrow \{h:X_H\rightarrow X_H\,|\,h\text{ a covering transformation}\}\]
  \[g\mapsto h_g:X_H\rightarrow X_H\]
\end{theorem}

\begin{theorem}
  \[Aut(X_H)\cong N(H)/H\]
\end{theorem}
\begin{definition}[Normal cover, universal cover and the deck group]
  If $H\triangleleft G$, then $X_H$ is said to be a \textit{normal cover}. If $H=1$, 
  then $X_H$ is simply connected, and is denoted $\tilde{X}$, called the \textit{universal
  cover}. We call $Aut(\tilde{X})$ the \textit{deck group}. \\

  \noindent 
  If such an $\tilde{X}$ exists, then \[Aut(\tilde{X})\cong G\text{ and }X_H=\tilde{X}/H\] In
  particular, $X=\tilde{X}/G$.
\end{definition}

\section{Free groups}
\begin{definition}
  A group $F$ is free on $S\subseteq F$ if for every group $G$ and map $\phi:S\rightarrow G$ there 
  exists a unique homomorphism $f:F\rightarrow G$ such that $f|_S=\phi$. $S$ is said 
  to be a \textit{basis} for $F$.
\end{definition}

  Given $S$, we construct $F_S$ as follows,
  \begin{enumerate}
    \item $S$ is a \textit{set}, thought of as a collection of symbols.
    \item We extend $S$ to $S^{\pm}=S\cup S^-$, where $S^-=\{s^{-1}\mid s\in S\}$. 
    \item Define $(S^{\pm})^*=\bigcup_{n\in\mathbb{N}}\{w:n\rightarrow S^{\pm}\}$.
    An element
    of $(S^{\pm})^*$ is a \textit{word}.
    \item There is a special word $\epsilon:0\rightarrow S^{\pm}$, the unique string of length 0.
  \end{enumerate}

\begin{definition}
  A word $w$ is \textit{reduced} if $w(i)=s\implies w(i-1)\neq s^{-1}=\neq w(i+1)$ for all $s\in S$, and also $w(i)=s^{-1}\implies w(i-1)\neq s\neq w(i+1)$ for all $s^{-1}\in S$.
\end{definition}

We write $w\rightarrow w'$ is $w'$ is obtained from $w$ by deleting an adjacent pair of $ss^{-1}$
or $s^{-1}s$ from $w$.

\begin{definition}
  $w$ reduces to $w'$ if there exists a sequence
  \[w=w_0\rightarrow w_1\rightarrow w_2\rightarrow\ldots\rightarrow w_k=w'\]
  where in this case, we write $w\implies w'$.
\end{definition}

\begin{lemma}
  If $w\implies w'$ and $w\implies w''$, then $w'=w''$.
\end{lemma}

\begin{definition}
  $F_S=\{\text{reduced words in }(S^\pm)^*\}$
\end{definition}

\begin{proposition}
  $F_S$ is a group, where $e=\epsilon$ is the empty string, $w^-1$ is $w$ in reverse, and swapping the signs of each letter. $w_1\cdot w_2$ is the unique word that $w_1w_2$ reduces 
  to.
\end{proposition}

\noindent\underline{Universal property} \\ 
Identify $s\in S$ with the string $s:1\rightarrow S^\pm$, $s(1)=s$. Given 
$\phi:S\rightarrow G$ (group), extend this to a homomorphism,
\[f(e)=e\]\[f(s)=\phi(s), f(s^{-1})=f(s)^{-1}\] Now define $f(w)=f(w(0))\cdot f(w(1))\cdot
f(w(2))\cdot\ldots\cdot f(w(n-1))$ when $w:n\rightarrow S^\pm$. It can be checked that 
this indeed defines a homomorphism.

\begin{definition}
  $w$ and $w'$ are homotopic if there exists $w=w_0,w_1,\ldots,w_n=w'$ such that either 
  $w_i\rightarrow w_{i+1}$ or $w_i\leftarrow w_{i+1}$ for all $i$.
\end{definition}

\begin{theorem}
  If $w$ and $w'$ are homotopic and reduced then $w=w'$.
\end{theorem}
Alternatively we could have also defined $F_S$ to be the set of homotopy classes of words.
The real important thing is that reduced representative is \textit{unique}.

\begin{definition}
  A presentation is a pair $\langle S\mid R\rangle$, where $S$ is a set and $R$ is 
  a collection of words in $S^\pm$. $\langle S\mid R\rangle$ is a presentation of $G$ if 
  $G\cong F_S/\langle\langle R'\rangle\rangle$ where $R'=\{\text{reduced words equivalent to elements in $R$}\}$. Recall that $\langle\langle R'\rangle\rangle$ is the smallest 
  normal subgroup containing $R'$.
\end{definition}

\begin{theorem}
  Let $X$ be a connected graph, then $\pi_1(X)$ is free.
\end{theorem}

\begin{definition}
  A \textit{graph} is a 1-dimensional CW-complex.
\end{definition}

\begin{corollary}
  Subgroups of free groups are free.
\end{corollary}

\begin{corollary}
  Every group is a quotient of a free group by a free group.
\end{corollary}

\begin{definition}
  A group $G$ is finitely generated if $\exists f:F_S\rightarrow G, |S|<\infty$.
\end{definition}

\begin{definition}
  A \textit{forest} is a graph with no embedded $S^1$'s (1-complex homeomorphic to $S^1$).
  A \textit{tree} is a connected forest. A forest $F\subseteq X$ is \textit{maximal} if for any 
  $F\subsetneq Y\subseteq X$, $Y$ is not a forest.
\end{definition}

\begin{proposition}
  Any graph $X$ contains a maximal forest, and if $X$ is connected, $F$ is free.
\end{proposition}
\end{document}

